%\section{Loops and lists}

\begin{Problem}{\textbf{Multiply by five}} \label{prob21}
%\addcontentsline{toc}{section}{Exercise 2.1: Multiply by five   - \texttt{multiplication.py}}

\noindent Write a code printing out $5\cdot 1$, $5\cdot 2$, ..., $5\cdot 10$, using either
a \pythoninline{for} or a \pythoninline{while} loop.

Filename: $\texttt{multiplication.py}$
\end{Problem}


\begin{Problem}{\textbf{Multiplication table}}\label{prob22}
%\addcontentsline{toc}{section}{Exercise 2.2: Multiplication table - \texttt{mult\_table.py}}

\noindent Write a new code based on the one from Problem \ref{prob21}. This code should print the
whole miltiplication table from $1\cdot 1$ to $10 \cdot 10$.

\emph{Hint: You may want to consider using one loop inside another.}

Filename: $\texttt{mult\_table.py}$
\end{Problem}

\begin{Problem}{\textbf{Stirling's approximation}}\label{prob23}
%\addcontentsline{toc}{section}{Exercise 2.3: Stirling's approximation - \texttt{stirling.py}}

\noindent Stirling's approximation can be written $\ln (x!) \approx x\ln x - x$. This is a 
good approximation for large $x$. Write out a nicely formatted table of integer
$x$ values, the actual value of $\ln (x!)$, and Stirling's approximation to $\ln (x!)$.

Filename: $\texttt{stirling.py}$
\end{Problem}
	
	
\begin{Problem}{\textbf{Errors in summation}}\label{prob24}
%\addcontentsline{toc}{section}{Exercise 2.4: Errors in summation  - \texttt{sum\_for.py}}

\noindent The following code is supposed to compute the sum $s = \sum_{k = 1}^{M}\frac{1}{(2k)^2}$, for $M=3$.
\begin{python}
s = 0 
M = 3

for i in range(M):
    s += 1/2*k**2

print(s)
\end{python}
The program has three errors and therefore does not work. Find the three errors and write a correct program. Put comments in your program to indicate what the mistakes were.

\emph{Hint:} There are two basic ways to find errors in a program:
\begin{enumerate}
    \item read the program carefully and think about the consequences of each statement,
    \item print out inter mediate results and compare with hand calculations.
\end{enumerate}
1. 
First, try method 1 and find as many errors as you can. Thereafter, try method 2 for M D 3 and compare the evolution of s with your own hand calculations.

Lastly, write a similar code evaluating the same sum using a \pythoninline{while}
loop. Check that the two loops compute the same answer.

Filename: $\texttt{sum\_for.py}$
\end{Problem}
\newpage	
	
\begin{Problem}{\textbf{Binomial coefficient}}\label{prob25}
%\addcontentsline{toc}{section}{Exercise 2.5: Binomial coefficient - \texttt{binomial.py}}

\noindent The binomial coefficient is indexed by two integers $n$ and $k$ and is written
$\binom{n}{k}$. It is given by the formula
\begin{equation}\label{eq: binomial_factorial}
\binom{n}{k} = \frac{n!}{k!(n-k)!}.
\end{equation}
We can write this out, and get
\begin{equation}\label{eq: binomial_product}
\binom{n}{k} = \prod_{j = 1}^{n - k}\frac{k + j}{j}.
\end{equation}

Use Eq.~(\ref{eq: binomial_product}) and a \pythoninline{for} loop to find the binomial coefficient
for $n = 14$ and $k = 3$. Compute the same value using Eq.~(\ref{eq: binomial_factorial})
and check that the results are correct.

\emph{Hint: The $\prod$ sign is a product sign. Thus
$\prod_{j = 1}^{n - k}\frac{k + j}{j} = \frac{k + 1}{1}\frac{k + 2}{2}
\dots\frac{k + (n - k)}{(n - k)}$. When checking the result you will need
\pythoninline{math.factorial}.} 

Filename: $\texttt{binomial.py}$
\end{Problem}	
	
	
	

\begin{Problem}{\textbf{Table showing population growth}} \label{population_table}

\noindent Consider again the bacterial colony from Problem \ref{prob12}. Let us study the number
of individuals for $n + 1$ uniformly spaced $t$ values throughout the interval
$[0, 48]$. Set $n = 12$. First store the $t$ and $N$ values in two lists \pythoninline{t}
and \pythoninline{N}. Thereafter, write out a nicely formatted table of $t$ and $N$
values by traversing the two lists with a (separate)  \pythoninline{for} loop. 

Filename: \texttt{population\_table.py}	
\end{Problem}	
	
\begin{Problem}{\textbf{Nested list}}\label{population_table2}

\paragraph{a)} Compute two lists of $t$ and $N$ values as explained in Problem
\ref{population_table}. Store the two lists in a new nested list \pythoninline{tN1} such that
\pythoninline{tN1[0]} is the list containing $t$-values and \pythoninline{tN[1]} correspond to the list containing $N$-values. Write out a table
with $t$ and $N$ values in two columns by looping over the data in the \pythoninline{tN1}
list. Each $t$ and $N$ value should be written in the table as integers. 

\paragraph{b)} Make a nested list \pythoninline{tN2} where \pythoninline{tN2[i]} contains the $i$-th element of both the $t$-list and the $N$-list. Loop over the \pythoninline{tN2} list and write out the $t$ and $N$
values in the table as integers. 


Filename: $\texttt{population\_table2.py}$	
\end{Problem}

\begin{Problem}{\textbf{Calculate Cesaro mean}}

\noindent Let $(a_n)_{n=1}^\infty$ be a sequence of numbers, $s_k=\sum_{n=0}^k a_n=a_0+\dots,+a_k$,
and
\begin{equation*}
    S_N = \frac{1}{N-1}\sum_{k=0}^{N-1} s_k.
\end{equation*}
Let $(a_n)_{n=1}^\infty$ be the sequence with $a_n=(-1)^n$.
Calculate $S_N$ for \newline $N=1, 2, 3, 4, 5, 10, 50$ and print the results in a table.

Filename: $\texttt{cesaro\_mean.py}$
\end{Problem}

\begin{Problem}{\textbf{Catalan numbers}}

\noindent A number on the form
\begin{equation*}
    C_n=\frac{1}{n+1}\binom{2n}{n}=\frac{(2n)!}{(n+1)!n!}
\end{equation*}
is called a \emph{Catalan number}. Compute and print the first 10 Catalan numbers.
    
Filename: $\texttt{catalan.py}$
\end{Problem}

\begin{Problem}{\textbf{Molar Mass of Alkanes}}

\noindent
Alkanes are saturated hydrocarbons with the chemical formula $\mathrm{C_nH_{2n+2}}$. If there are $n$ Carbon atoms in the alkane, there will be $m = 2n+2$ Hydrogen atoms. The molar mass of the hydrocarbon is $M_{\mathrm{C_n H_{m}}} = n M_{\mathrm{C}} + m M_{\mathrm{H}} $, where $M_C$ is the molar mass of Carbon and $M_H$ is the molar mass of Hydrogen. 

Use a \pythoninline{for}-loop or a \pythoninline{while}-loop to compute and print out the molar mass of the alkanes with two through nine Carbon atoms ($n \in [2, 9]$). The output should specify the chemical formula of the alkane as well as the molar mass. An example on how the formatted output should look like for $n = 2$ is given below.
\begin{lstlisting}
M(C2H6) =  30.069 g/mol
\end{lstlisting}
You can set the molar masses of the atoms to be $M_{\mathrm{C}} = 12.011 \ \mathrm{g/mol}$ and $M_{\mathrm{H}}  = 1.0079 \ \mathrm{g/mol}$
    
Filename: $\texttt{alkane.py}$
\end{Problem}


\begin{Problem}{\textbf{Matrix elements}}

\begin{center}
\emph{This exercise involves no programming. \newline The answers should be written in a text file called $\texttt{matrix.txt}$}    
\end{center}

\noindent Consider a two dimensional $3 \times 3$ matrix 

\begin{equation*}
   A = \begin{pmatrix} 
        a_{1 1} & a_{1 2} & a_{1 3} \\
        a_{2 1} & a_{2 2} & a_{2 3} \\
        a_{3 1} & a_{3 2} & a_{3 3} 
    \end{pmatrix} .
\end{equation*}

In Python, the matrix $A$ can be represented as a nested list \pythoninline{A}, either as a list of rows or a list of columns. Find the indices \pythoninline{i, j} of the Python list \pythoninline{A} such that \pythoninline{A[i][j]} = $a_{1 1}$ and the indices \pythoninline{k, l}  such that \pythoninline{A[k][l]} = $a_{3 2}$ for the two following cases:

\paragraph{a)}
When $A$ is represented as a list of rows. This means that \pythoninline{A} contains three lists, where each list corresponds to a row in $A$.

\paragraph{b)}
When $A$ is represented as a list of columns. This means that each element in \pythoninline{A} contains a list with the elements of a column in $A$.

    
Filename: $\texttt{matrix.txt}$
\end{Problem}